\bf Tuning parameter selection for a penalized estimator of species richness

### 2019-08-28 rre_manuscript/data_analysis_extended.tex

@@

The richness estimates from each method are displayed in Figure \ref{fig:data_analysis} and Table \ref{tab:data_analysis}.  These estimates are high as 160 times the observed richness, suggesting that the negative binomial model may be a poor fit to this data set.  As a result, applying our methods to real data required some manual tuning of $\lambdagrid$ and the search grid for $C$.

Despite the estimates being significantly more inflated than in our simulations, many of the trends observed in simulations hold true here.  Method 0 produces the highest $\widehat{C}$ for 2009 and 2010 and it was second highest to Method 2 in 2011.  Method 1 consistently estimated $\widehat{C}$ to be at or near the observed richness $c$.  These are exactly the tendencies these methods showed when we generate simulations from a correctly specified model (see Figure \ref{fig:tuning_sim_1}).

We model species richness because the observed richness is an underestimate of the truth.  The original motivation for penalization is that the unpenalized MLE estimate can be much larger than the truth.  In data analysis we don't know the truth, but we would expect an ideal method to be situated between these two extremes.  Only the methods making use of a goodness of fit metric (3 and 4) are between Method 0 and the observed richness in all samples.  As a result we conclude Methods 3 and 4 show the greatest promise on real data containing replicates.

\noindent\large{\textbf{Note from Alex:  I've given two versions of the table, the first is much bigger and probably more suited to being a supplement.  The second is more compact and could be crammed into a corner of the paper if needed.  Happy to make revisions to tables and figures if that's helpful. }}

\begin{table}[ht]
\caption{Diversity estimates from the Lake Champlain data analysis from 2009 ($r = 8$), 2010 ($r = 6$) and 2011 ($r = 6$) from our proposed methods. 
\label{tab:data_analysis}}
\centering
\begin{tabular}{llcccc}
  \toprule
Year & Method & $\widehat{C}$ & $\widetilde{\lambda}$ & $\widehat{\alpha}$ & $\widehat{\delta}$ \\ 
  \midrule
2009 & {[0]} Unpenalized MLE & 73,404 & \textemdash & 0.00088 & 0.00180 \\ 
  \multirow{5}{*} & {[1]} Minimum Variance & 593 & 700 & 0.27720 & 0.00342 \\ 
  \multirow{5}{*} & {[2]} CV Likelihood & 25,930 & 220 & 0.00516 & 0.00142 \\ 
  \multirow{5}{*} & {[3]} Goodness of fit & 20,160 & 550 & 0.00323 & 0.00174 \\ 
  \multirow{5}{*} & {[4]} CV G.O.F. & 39,997 & 100 & 0.00578 & 0.00253 \\ 
  2010 & {[0]} Unpenalized MLE & 47,631 & \textemdash & 0.00185 & 0.00253 \\ 
  \multirow{5}{*} & {[1]} Minimum Variance & 572 & 500 & 0.62668 & 0.00724 \\ 
  \multirow{5}{*} & {[2]} CV Likelihood & 24,799 & 0 & 0.00379 & 0.00270 \\ 
  \multirow{5}{*} & {[3]} Goodness of fit & 13,156 & 220 & 0.00685 & 0.00258 \\ 
  \multirow{5}{*} & {[4]} CV G.O.F. & 47,098 & 0 & 0.00208 & 0.00277 \\ 
  2011 & {[0]} Unpenalized MLE & 57,686 & \textemdash & 0.00161 & 0.00140 \\ 
  \multirow{5}{*} & {[1]} Minimum Variance & 718 & 500 & 0.40112 & 0.00355 \\ 
  \multirow{5}{*} & {[2]} CV Likelihood & 118,547 & 0 & 0.00122 & 0.00193 \\ 
  \multirow{5}{*} & {[3]} Goodness of fit & 40,040 & 230 & 0.00231 & 0.00137 \\ 
  \multirow{5}{*} & {[4]} CV G.O.F. & 36,395 & 5 & 0.00358 & 0.00178 \\ 
   \bottomrule
\end{tabular}
\end{table}

\begin{table}[ht]
\caption{Diversity estimates from the Lake Champlain data analysis from 2009 ($r = 8$), 2010 ($r = 6$) and 2011 ($r = 6$) from our proposed methods. 
\label{tab:data_analysis_compact}}
\centering
\tiny
\begin{tabular}{lllllllllllll}
  \toprule
      & \multicolumn{4}{c}{2009} & \multicolumn{4}{c}{2010} & \multicolumn{4}{c}{2011} \\ \cmidrule(lr){2-5}\cmidrule(lr){6-9}\cmidrule(lr){10-13}
Method & $\widehat{C}$ & $\widetilde{\lambda}$ & $\widehat{\alpha}$ & $\widehat{\delta}$ & $\widehat{C}$ & $\widetilde{\lambda}$ & $\widehat{\alpha}$ & $\widehat{\delta}$ & $\widehat{C}$ & $\widetilde{\lambda}$ & $\widehat{\alpha}$ & $\widehat{\delta}$ \\ 
  \midrule
{[0]} Unpenalized MLE & 73,404 & \textemdash & 0.00088 & 0.00180 & 47,631 & \textemdash & 0.00185 & 0.00253 & 57,686 & \textemdash & 0.00161 & 0.00140 \\ 
  {[1]} Minimum Variance & 593 & 700 & 0.27720 & 0.00342 & 572 & 500 & 0.62668 & 0.00724 & 718 & 500 & 0.40112 & 0.00355 \\ 
  {[2]} CV Likelihood & 25,930 & 220 & 0.00516 & 0.00142 & 24,799 & 0 & 0.00379 & 0.00270 & 118,547 & 0 & 0.00122 & 0.00193 \\ 
  {[3]} Goodness of fit & 20,160 & 550 & 0.00323 & 0.00174 & 13,156 & 220 & 0.00685 & 0.00258 & 40,040 & 230 & 0.00231 & 0.00137 \\ 
  {[4]} CV G.O.F. & 39,997 & 100 & 0.00578 & 0.00253 & 47,098 & 0 & 0.00208 & 0.00277 & 36,395 & 5 & 0.00358 & 0.00178 \\ 
   \bottomrule
\end{tabular}
\end{table}

@@

\bibliographystyle{biorefs}
\bibliography{refs,amys-papers-library}

>>>>

\noindent \textbf{Note from Alex:}\textit{ Included is my data analysis draft, a figure, a table and some blue comments which may or may not be useful in discussion.  I did the table two ways - the same information is contained in both versions (The \texttt{mutlirow} package is needed for the first one.)  As always happy to edit text, figures or tables if that's helpful.}

@@

The richness estimates from each method are displayed in Figure \ref{fig:data_analysis}.  Method 0 produces the highest estimate for 2009 ($\widehat{C}_{[0]} = 73,404$) and 2010 ($\widehat{C}_{[0]}$ = 47,631) and it was second highest to Method 2 in 2011 ($\widehat{C}_{[2]} = 118,547$).  Method 1 consistently estimated $\widehat{C}$ to be at or near the observed richness, $C_{[1]} = 584$ in 2009,  $572$ in 2010 and $718$ in 2011.  Methods 3 and 4, which both use goodness of fit, were between these two extremes.  These results mirror what we observed in simulations, where Methods 0 and 2 were the highest estimates and Method 1 was the lowest (see Figure \ref{fig:tuning_sim_1}).

These estimates are up to 160 times the observed richness, suggesting that a majority of the species were unobserved in each sample.  The relatively inflated estimates suggest that these data may be poorly fit by the negative binomial model.  \\

\noindent \textcolor{blue}{\textbf{Notes for discussion:}
The reason we model species richness is that the observed richness is an underestimate of the actual richness.  The motivation for penalization is that the unpenalized MLE tends to be much larger than the actual richness.  These results show that only the methods making use of a goodness of fit metric (3 and 4) were consistently between Method 0 and the observed richness in all samples.  We conclude Methods 3 and 4 show the greatest promise on real data containing replicates.}

\textcolor{blue}{
In order to apply our methods to real data, some manual tuning of the optimization search grids ($\lambdagrid$ and the search grid for $C$) was required.}

\normalsize

\begin{table}[ht]
\caption{Diversity estimates from the Lake Champlain data analysis from 2009 ($r = 8$), 2010 ($r = 6$) and 2011 ($r = 6$) from our proposed methods. 
\label{tab:data_analysis}}
\centering
\begin{tabular}{llcccc}
\toprule
Year & Method & $\widehat{C}$ & $\widetilde{\lambda}$ & $\widehat{\alpha}$ & $\widehat{\delta}$ \\ 
  \midrule
2009 & {[0]} Unpenalized MLE & 73,404 & \textemdash & 0.00088 & 0.00180 \\ 
  \multirow{5}{*} & {[1]} Minimum Variance & 584 & 700 & 0.28863 & 0.00326 \\ 
  \multirow{5}{*} & {[2]} CV Likelihood & 25,930 & 220 & 0.00516 & 0.00142 \\ 
  \multirow{5}{*} & {[3]} Goodness of fit & 20,160 & 550 & 0.00323 & 0.00174 \\ 
  \multirow{5}{*} & {[4]} CV G.O.F. & 39,997 & 100 & 0.00578 & 0.00253 \\ 
  2010 & {[0]} Unpenalized MLE & 47,631 & \textemdash & 0.00185 & 0.00253 \\ 
  \multirow{5}{*} & {[1]} Minimum Variance & 572 & 500 & 0.62668 & 0.00724 \\ 
  \multirow{5}{*} & {[2]} CV Likelihood & 24,799 & 0 & 0.00379 & 0.00270 \\ 
  \multirow{5}{*} & {[3]} Goodness of fit & 13,156 & 220 & 0.00685 & 0.00258 \\ 
  \multirow{5}{*} & {[4]} CV G.O.F. & 47,098 & 0 & 0.00208 & 0.00277 \\ 
  2011 & {[0]} Unpenalized MLE & 57,686 & \textemdash & 0.00161 & 0.00140 \\ 
  \multirow{5}{*} & {[1]} Minimum Variance & 718 & 500 & 0.40112 & 0.00355 \\ 
  \multirow{5}{*} & {[2]} CV Likelihood & 118,547 & 0 & 0.00122 & 0.00193 \\ 
  \multirow{5}{*} & {[3]} Goodness of fit & 40,040 & 230 & 0.00231 & 0.00137 \\ 
  \multirow{5}{*} & {[4]} CV G.O.F. & 36,395 & 5 & 0.00358 & 0.00178 \\ 
   \bottomrule
\end{tabular}
\end{table}

\begin{table}[ht]
\caption{Diversity estimates from the Lake Champlain data analysis from 2009 ($r = 8$), 2010 ($r = 6$) and 2011 ($r = 6$) from our proposed methods. 
\label{tab:data_analysis_compact}}
\centering
\tiny
\begin{tabular}{lllllllllllll}
  \toprule
      & \multicolumn{4}{c}{2009} & \multicolumn{4}{c}{2010} & \multicolumn{4}{c}{2011} \\ \cmidrule(lr){2-5}\cmidrule(lr){6-9}\cmidrule(lr){10-13}
Method & $\widehat{C}$ & $\widetilde{\lambda}$ & $\widehat{\alpha}$ & $\widehat{\delta}$ & $\widehat{C}$ & $\widetilde{\lambda}$ & $\widehat{\alpha}$ & $\widehat{\delta}$ & $\widehat{C}$ & $\widetilde{\lambda}$ & $\widehat{\alpha}$ & $\widehat{\delta}$ \\ 
  \midrule
{[0]} Unpenalized MLE & 73,404 & \textemdash & 0.00088 & 0.00180 & 47,631 & \textemdash & 0.00185 & 0.00253 & 57,686 & \textemdash & 0.00161 & 0.00140 \\ 
  {[1]} Minimum Variance & 584 & 700 & 0.28863 & 0.00326 & 572 & 500 & 0.62668 & 0.00724 & 718 & 500 & 0.40112 & 0.00355 \\ 
  {[2]} CV Likelihood & 25,930 & 220 & 0.00516 & 0.00142 & 24,799 & 0 & 0.00379 & 0.00270 & 118,547 & 0 & 0.00122 & 0.00193 \\ 
  {[3]} Goodness of fit & 20,160 & 550 & 0.00323 & 0.00174 & 13,156 & 220 & 0.00685 & 0.00258 & 40,040 & 230 & 0.00231 & 0.00137 \\ 
  {[4]} CV G.O.F. & 39,997 & 100 & 0.00578 & 0.00253 & 47,098 & 0 & 0.00208 & 0.00277 & 36,395 & 5 & 0.00358 & 0.00178 \\ 
   \bottomrule
\end{tabular}
\end{table}

@@

%\bibliographystyle{biorefs}
%\bibliography{refs,amys-papers-library}
